---
title: "Preliminary data"
author: "Arpan Neupane"
date: "24/06/2020"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose
To demonstrate the issue my project is attempting to address. 
dataframe has been saved as rdata

```{r eval=T}
library(tidyverse)
path <- paste(getwd(),"datafiles","prelim_data",sep = "/")
my_data <- readRDS(file = path)
glimpse(my_data)
```

```{r }
glimpse(my_data)
```

## Top 20 authors
The author ordering hasn't been kept. For my exploratory purposes, all authors get an equal chance to be represented, even the middle ones.

```{r}

t_20 <- my_data %>% na.omit %>%filter(authorlist != "NA NA") %>% 
  group_by(authorlist,dataset) %>% summarize(total_pub = n()) %>% arrange(desc(total_pub)) %>% select(authorlist) %>% 
  pull %>% head(20)
t_20
```

Florent Ginhoux is a very well known name in macrophage and DC biology. So far so good.
Note also that Jing Wang and Wei Wang pop up twice. Maybe they work on different fields?

Let's visualize the first preliminary plot.

#First plot
```{r}
initial_plot <- my_data %>% filter(authorlist %in% t_20) %>% group_by(authorlist,journal,dataset) %>% summarize(total_pub = n()) %>% 
  arrange(desc(total_pub)) %>% 
  ggplot(aes(x = reorder(authorlist,total_pub), y= total_pub))+
  geom_col(aes(fill = dataset),position = "fill")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90))+
  xlab("Top 20 authors in the downloaded dataset \nBy number of publications\nAuthorname:Last First")+
  ylab("proportion in\neach dataset")+
  labs(title="Author contribution to each dataset",subtitle = "PMC data from 2018-2020 for select search terms",
        caption = "Data (~5000) pertaining to search terms (dataset) were downloaded from PMC on 2020-06-23.\n
        Does not fully reflect all publications from 2018-2020")

initial_plot
```

This shows already that there exists a lot of "crossing over" of authors between different dataset. 
Thus, even though we might have asked for paper by Florent Ginhoux, because we are interested in DCs, wouldn't it be worthwhile to also get to see other researchers that frequently work with Florent on DCs? 
The present solution is click and scroll --> rinse and repeat until you find everyone you need. Cumbersome!

## Did the search term really capture all the essence of the paper?

This is an extremly rudementary and cude approach to contextualize the search results. 

We will parse through the title+abstract for some key terms, which still relate to the query term, and cross examine "crossing over".

```{r}
search_terms <- c("b cells|b-cells|b cell|b-cell","t-cells|t cells|t cell|t-cell","dendritic|dc|dcs|dendritic cells","macrophages|macrophage",
                  "neutrophil|neutrophils","lymph node|LN|lymph nodes")

second_plot <- my_data %>% filter(authorlist %in% t_20) %>% mutate(combine_both = paste0(abstract,articletitle)) %>% 
  mutate(title_contains= case_when(
  grepl(search_terms[1],combine_both,ignore.case = T)~"B Cell related",
  grepl(search_terms[2],combine_both, ignore.case = T)~"T Cell related",
  grepl(search_terms[3],combine_both, ignore.case = T)~"Dendritic Cell related",
  grepl(search_terms[4],combine_both, ignore.case = T)~"Macrophage related",
  grepl(search_terms[5],combine_both, ignore.case = T)~"Neutrophil related",
  grepl(search_terms[6],combine_both, ignore.case = T)~"Lymph Node related")
  ) %>% select(-combine_both) %>% group_by(authorlist, dataset,title_contains) %>% 
  summarize(top_pub = n()) %>% na.omit %>% 
  ggplot(aes(x = reorder(authorlist, desc(top_pub)), y = top_pub))+
  geom_col(aes(fill = dataset), position = "fill")+
  facet_wrap(~title_contains)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90))+
  labs(title = "Additional layer of classification of publications\nBy top 20 authors",
       subtitle = "PMC data from 2018-2020 for select search terms",
       caption = "Title and abstract were jointly searched for presence of select keywords.\n
       This is not an extensive characterization, rather an attempt to demonstrate \"crossing-over\" of authors into\n
       different subfields")+
    xlab("Top 20 authors")+ylab("propotion")
second_plot
```

Even with such a rudemenatry approach, one can already start to make interesting observations. 
For example: Titles+abstracts that contained B cell related terms were present in datasets prepared
by searching fo b cells, neutrophils, dendritic cells, t cells etc. 

## Conclusion:

Further analysis (maybe some "light reading" of the abstracts) will need to be done to confirm that 
the patterns we see in the plot above do indeed make sense. 
More importantly, the plots above present the issue that scientif research is highly diverse, with 
frequent "crossing-over" of authors into different subfields (the example was from immunological perspective).
Using text based search method can be tedious. A graphical and interactive approach to visualizing PMC search results
may expedite researching a topic/finding prominent authors and labs/collaboration networks etc.


